\begin{thebibliography}{10}

\bibitem{akiyama_petho2002:canonical_number_systems}
S.~Akiyama and A.~Peth{\H{o}}.
\newblock On canonical number systems.
\newblock {\em Theoret. Comput. Sci.}, 270(1-2):921--933, 2002.

\bibitem{Bugeaud:2002}
Y.~Bugeaud, G.~Hanrot, and M.~Mignotte.
\newblock Sur l'\'equation diophantienne {$(x^n-1)/(x-1)=y^q$}. {III}.
\newblock {\em Proc. London Math. Soc. (3)}, 84(1):59--78, 2002.

\bibitem{Chao:1965}
K.~Chao.
\newblock On the {D}iophantine equation {$x^{2}=y^{n}+1,\,xy\not=0$}.
\newblock {\em Sci. Sinica}, 14:457--460, 1965.

\bibitem{cobham1972:uniform_tag_sequences}
A.~Cobham.
\newblock Uniform tag sequences.
\newblock {\em Math. Systems Theory}, 6:164--192, 1972.

\bibitem{gilbert81:_radix}
W.~J. Gilbert.
\newblock Radix representations of quadratic fields.
\newblock {\em J. Math. Anal. Appl.}, 83(1):264--274, 1981.

\bibitem{hansel_safer2003:vers_un_theoreme}
G.~Hansel and T.~Safer.
\newblock Vers un th\'eor\`eme de {C}obham pour les entiers de {G}auss.
\newblock {\em Bull. Belg. Math. Soc. Simon Stevin}, 10(suppl.):723--735, 2003.

\bibitem{katai_kovacs1980:kanonische_zahlensysteme_in}
I.~K{\'a}tai and B.~Kov{\'a}cs.
\newblock Kanonische {Z}ahlensysteme in der {T}heorie der quadratischen
  algebraischen {Z}ahlen.
\newblock {\em Acta Sci. Math. (Szeged)}, 42(1-2):99--107, 1980.

\bibitem{katai_kovacs1981:canonical_number_systems}
I.~K{\'a}tai and B.~Kov{\'a}cs.
\newblock Canonical number systems in imaginary quadratic fields.
\newblock {\em Acta Math. Acad. Sci. Hungar.}, 37(1-3):159--164, 1981.

\bibitem{katai_szabo1975:canonical_number_systems}
I.~K{\'a}tai and J.~Szab{\'o}.
\newblock Canonical number systems for complex integers.
\newblock {\em Acta Sci. Math. (Szeged)}, 37(3-4):255--260, 1975.

\bibitem{knuth1981:art_computer_programming}
D.~E. Knuth.
\newblock {\em The art of computer programming. {V}ol. 2}.
\newblock Addison-Wesley Publishing Co., Reading, Mass., second edition, 1981.
\newblock Seminumerical algorithms, Addison-Wesley Series in Computer Science
  and Information Processing.

\bibitem{kovacs1981:canonical_number_systems}
B.~Kov{\'a}cs.
\newblock Canonical number systems in algebraic number fields.
\newblock {\em Acta Math. Acad. Sci. Hungar.}, 37(4):405--407, 1981.

\bibitem{kovacs_petho1991:number_systems_in}
B.~Kov{\'a}cs and A.~Peth{\H{o}}.
\newblock Number systems in integral domains, especially in orders of algebraic
  number fields.
\newblock {\em Acta Sci. Math. (Szeged)}, 55(3-4):287--299, 1991.

\bibitem{Luca:2009}
F.~Luca, S.~Tengely, and A.~Togb{\'e}.
\newblock On the {D}iophantine equation {$x^2+C=4y^n$}.
\newblock {\em Ann. Sci. Math. Qu\'ebec}, 33(2):171--184, 2009.

\bibitem{Mihailescu:2004}
P.~Mih{\v{a}}ilescu.
\newblock Primary cyclotomic units and a proof of {C}atalan's conjecture.
\newblock {\em J. Reine Angew. Math.}, 572:167--195, 2004.

\bibitem{pethoe1991:polynomial_transformation_and}
A.~Peth{\H o}.
\newblock On a polynomial transformation and its application to the
  construction of a public key cryptosystem.
\newblock In {\em Computational number theory ({D}ebrecen, 1989)}, pages
  31--43. de Gruyter, Berlin, 1991.

\bibitem{Schinzel:1976}
A.~Schinzel and R.~Tijdeman.
\newblock On the equation {$y^{m}=P(x)$}.
\newblock {\em Acta Arith.}, 31(2):199--204, 1976.

\bibitem{Shorey:Dioph}
T.~N. Shorey and R.~Tijdeman.
\newblock {\em Exponential {D}iophantine equations}, volume~87 of {\em
  Cambridge Tracts in Mathematics}.
\newblock Cambridge University Press, Cambridge, 1986.

\end{thebibliography}
\documentclass{amsart}
\usepackage{amsmath,amsfonts,amssymb,latexsym}
\usepackage{graphicx}
\usepackage{fancyhdr}
\usepackage{multirow}

%\usepackage{a4wide}

\usepackage{cite}

\usepackage{hyperref}

\hypersetup{%
pdftitle={},
pdfsubject={Mathematics},
pdfauthor={},
pdfkeywords={},
hyperindex=true,plainpages=false}

\newcommand{\R}{{\mathbb R}}
\newcommand{\Q}{{\mathbb Q}}
\newcommand{\C}{{\mathbb C}}
\newcommand{\N}{{\mathbb N}}
\newcommand{\Z}{{\mathbb Z}}
\renewcommand{\P}{{\mathbb P}}
\newcommand{\ord}{\mathrm{ord}}

\DeclareMathOperator{\Rad}{Rad}
\DeclareMathOperator{\Gal}{Gal}
\DeclareMathOperator{\Norm}{N}

\theoremstyle{plain}
\numberwithin{equation}{section}
\newtheorem{thm}{Theorem}[section]
\newtheorem{theorem}[thm]{Theorem}
\newtheorem{example}[thm]{Example}
\newtheorem{lemma}[thm]{Lemma}
\newtheorem{definition}[thm]{Definition}
\newtheorem{proposition}[thm]{Proposition}
\newtheorem{conjecture}[thm]{Conjecture}
\newtheorem{question}[thm]{Question}
\newtheorem{cor}[thm]{Corollary}
\newtheorem{con}[thm]{Conjecture}

\theoremstyle{remark}
\newtheorem{remark}{Remark}

\begin{document}

%\setcounter{page}{1}

\title[Infinite family of bases]
{An infinite family of multiplicatively independent bases of number systems in cyclotomic number fields}

\author[M. G. Madritsch]{Manfred G. Madritsch}
\address{Manfred Madritsch\newline
\noindent 1. Universit\'e de Lorraine, Institut Elie Cartan de Lorraine, UMR 7502, Vandoeuvre-l\`es-Nancy, F-54506, France;\newline
\noindent 2. CNRS, Institut Elie Cartan de Lorraine, UMR 7502, Vandoeuvre-l\`es-Nancy, F-54506, France}
\email{manfred.madritsch@univ-lorraine.fr}

\author{Volker Ziegler}
\address{Volker Ziegler\newline
\noindent Johann Radon Institute for Computational and Applied Mathematics (RICAM)\newline
\noindent Austrian Academy of Sciences\newline
\noindent Altenbergerstr. 69\newline
\noindent A-4040 Linz, Austria}
\email{volker.ziegler\char'100ricam.oeaw.ac.at}
\thanks{The second author was supported by the Austrian Science Fund (FWF) under the project P~24801-N26.}

\begin{abstract}
  Let $\zeta_k$ be a $k$-th primitive root of unity, $m\geq\phi(k)+1$ an
  integer and $\Phi_k(X)\in\Z[X]$ the $k$-th cyclotomic polynomial.
  In this paper we show that the pair $(-m+\zeta_k,\mathcal N)$ is a
  canonical number system, with $\mathcal
  N=\{0,1,\dots,|\Phi_k(m)|\}$.  Moreover we also discuss whether the
  two bases $-m+\zeta_k$ and $-n+\zeta_k$ are multiplicatively
  independent for positive integers $m$ and $n$ and $k$ fixed.
\end{abstract}

\subjclass[2010]{11A63,11D61,11D41} \keywords{Canonical number
  systems, Radix representations, Diophantine equations,
  Nagell-Ljunggren equation}

\maketitle

\section{Introduction}
Let $q\geq2$ be a positive integer. Then the pair
$(q,\{0,1,\ldots,q-1\})$ is a number system in the positive integers,
\textit{i.e.} every integer $n\in\N$ has a unique representation of
the form
$$n=\sum_{j\geq0}a_jq^j\quad(a_j\in\{0,1,\ldots,q-1\}).$$

Taking a negative integer $q\leq-2$ and $\{0,1,\ldots,\lvert
q\rvert-1\}$ as set of digits one gets a number system for all
integers $\Z$.  This concept of a number system was further extended
to the Gaussian integers by
Knuth~\cite{knuth1981:art_computer_programming}. He showed that the
pairs $(-1+i,\{0,1\})$ and $(-1-i,\{0,1\})$ are number systems in the
Gaussian integers, \textit{i.e.} every Gaussian integer $n\in\Z[i]$
has a unique representation of the form
$$n=\sum_{j\geq0}a_jb^j\quad(a_j\in\{0,1\}),$$
where $b=-1+i$ or $b=-1-i$. It was shown by K{\'a}tai and Szab{\'o}
\cite{katai_szabo1975:canonical_number_systems} that all possible
bases for the Gaussian integers are of the form $-m\pm i$ with $m$ a positive
integer. These results were extended independently by K\'atai and
Kov\'acs \cite{katai_kovacs1981:canonical_number_systems,
  katai_kovacs1980:kanonische_zahlensysteme_in} and
Gilbert~\cite{gilbert81:_radix}, who classified all quadratic
extensions.

A different point of view on the above systems is the following. Let
$P=p_dX^d+p_{d-1}X^{d-1}+\cdots+p_1X+p_0\in\Z[X]$ with $p_d=1$ and
$\mathcal{R}=\Z[X]/P(X)\Z[X]$. Then we call the pair $(P,\mathcal{N})$
with $\mathcal{N}=\{0,1,\ldots,\lvert p_0\rvert-1\}$ a canonical
number system if every element $\gamma\in\mathcal{R}$ has a unique
representation of the form
$$\gamma=\sum_{j\geq0}a_jX^j\quad(a_j\in\{0,1,\ldots,p_0\}).$$
If $P$ is irreducible and $\beta$ is one of its roots, then
$\mathcal{R}$ is isomorphic to $\Z[\beta]$. In this case we simply
write $(\beta,\mathcal{N})$ instead of $(P,\mathcal{N})$. By setting
$P=X+q$ or $P=X^2+2mX+(m^2+1)$ we obtain the canonical number systems
in the integers and Gaussian integers from above, respectively.

Kov\'acs \cite{kovacs1981:canonical_number_systems} proved that for
any algebraic number field $K$ and order $\mathcal R=\Z[\alpha]$ of
$K$ there exists $\beta$ such that $(\beta,\mathcal{N})$ is a
canonical number system for $\mathcal R$. Moreover he proved that if
$1\leq p_{d-1}\leq \cdots\leq p_1\leq p_0$, $p_0\geq2$ and if $P$ is
irreducible, then $(P,\mathcal{N})$ is a canonical number system in
$\mathcal{R}$. Peth{\H o}
\cite{pethoe1991:polynomial_transformation_and} weakened the
irreducibility condition by only assuming that no root of the
polynomial is a root of unity.

Kov\'acs and Peth{\H o} \cite{kovacs_petho1991:number_systems_in}
provided necessary and sufficient conditions on the pair
$(\beta,\mathcal{N})$ to be a number system in $\Z[\alpha]$. A decade
later Akiyama 
and Peth{\H o} \cite{akiyama_petho2002:canonical_number_systems}
significantly reduced the number of cases one has to check under the
additional assumption that
$$\sum_{i=1}^d\lvert p_i\rvert<p_0.$$

Let us denote by $\zeta_k$ some primitive $k$-th root of unity. Since
$+i$ and $-i$ are primitive fourth roots of unity, we can say that all
the bases for the Gaussian integers are of the form $-m+\zeta_4$, with
$m\geq 1$. Furthermore an easy computation shows that $-m+\zeta_3$ are
bases in $\Z[\zeta_3]$ for $m\geq1$. Thus our first result answers the
question, whether $-m+\zeta_k$ is a basis in $\Z[\zeta_k]$.

\begin{theorem}\label{th:bases}
  Let $k>2$ and $m$ be positive integers. If $m\geq\phi(k)+1$, then
  $\{-m+\zeta_k,\mathcal{N}\}$ is a canonical number system.
\end{theorem}

Our second result considers, whether these bases are multiplicatively
independent. We call two algebraic integers $\alpha$ and $\beta$
multiplicatively independent, if $\alpha^p=\beta^q$ has only the
solution $p=q=0$ over the integers. Considering bases in the Gaussian
integers Hansel and Safer \cite{hansel_safer2003:vers_un_theoreme}
proved that $-m+i$ and $-n+i$ are multiplicative independent for all
$m>n>0$. Their motivation was a version of Cobham's theorem
\cite{cobham1972:uniform_tag_sequences} for the Gaussian
integers. Cobham's theorem states that a set $E\subset\N$ is $m$- and
$n$-recognizable with $m$ and $n$ multiplicative independent integers
if and only if the set $E$ is a union of arithmetic progressions. This
means that e.g. except for arithmetic progressions we are not able to
deduce the base $3$ expansion form the one in base $2$.

In the present paper we want to generalize the result of Hansel and
Safer \cite{hansel_safer2003:vers_un_theoreme} on the multiplicative
independence to bases of $\Z[\zeta_k]$ of the form $-m+\zeta_k$ for
$k\geq3$.

\begin{theorem}\label{th:mult_independent}
  Let $k\geq3$ be a positive integer. Then the algebraic integers
  $-m+\zeta_k$ and $-n+\zeta_k$ are multiplicatively independent
  provided $m>n>C(k)$, where $C(k)$ is an effective computable
  constant depending on $k$.

  Moreover, if $k$ is a power of $2,3,5,6,7,11,13$, $17,19$ or $23$,
  then $-m+\zeta_k$ and $-n+\zeta_k$ are multiplicatively independent
  as long as $m>n>0$.
\end{theorem}

In order to prove Theorem~\ref{th:mult_independent} one has to show
that the Diophantine equation
\begin{equation}\label{eq:mult_independent}
(-m+\zeta_k)^p=(-n+\zeta_k)^q
\end{equation}
has no solution $(n,m,p,q)$, with $m>n>C(k)$ and $p,q$ are not both
zero. Taking norms in \eqref{eq:mult_independent} we obtain
$\Phi_k(m)^p=\Phi_k(n)^q$ and therefore the Diophantine equation
\eqref{eq:mult_independent} becomes
\begin{equation}\label{eq:Norm2}
\Phi_k(m)^p=\eta^q,
\end{equation}
where $\Phi_k$ is the $k$-th cyclotomic polynomial and
$\eta=\Phi_k(n)$. In case that $p$ and $q$ have
greatest common divisor $d>1$ we take $d$-th roots and obtain equation
\eqref{eq:Norm2} with $p$ and $q$ co-prime. Therefore we may assume
that $p$ and $q$ are coprime and we deduce that
$\eta$ is a $p$-th power, i.e.  we get an equation of the form
$\Phi_k(m)=y^q$ with $y=\eta^{1/p}$.   Note that since we assume
$m,n>C(k)$ the case $\Phi_k(m)=-y^q$ can be excluded. Indeed, we know
that $\Phi_k(m)>0$ for $m$ sufficiently large. In any case we obtain a
Diophantine equation of the form
\begin{equation}\label{eq:Norm}
\Phi_k(m)=y^q.
\end{equation}
Equation \eqref{eq:Norm} is closely related to the well-studied Nagell-Ljunggren equation
\begin{equation}\label{eq:Nagell}
\frac{x^k-1}{x-1}=y^q\qquad x,y>1,\; q\geq 2, \; k>2.
\end{equation}
If $k$ is a prime, then the Diophantine equations \eqref{eq:Norm} and \eqref{eq:Nagell} are identical and every solution to \eqref{eq:mult_independent} yields 
a solution to \eqref{eq:Nagell}. Therefore we conjecture: 

\begin{conjecture}
Let $k$ be an odd prime. Then $-m+\zeta_k$ and $-n+\zeta_k$ are multiplicatively independent provided $m>n>0$.
\end{conjecture}

Since it is widely believed that the Nagell-Ljunggren equation \eqref{eq:Nagell} has only the three solutions
\[\frac{3^5-1}{3-1}=11^2,\qquad \frac{7^4-1}{7-1}=20^2, \qquad \frac{18^3-1}{18-1}=7^3\]
this conjecture seems plausible. Note that none of these solutions to \eqref{eq:Nagell} implies a solution to \eqref{eq:mult_independent}. We guess that even 
more 
is true and believe that the following question has an affirmative answer.

\begin{question}
 Given $k>2$ are $-m+\zeta_k$ and $-n+\zeta_k$ multiplicatively independent provided $m>n>1$?
\end{question}

We excluded the case that $n=1$ since for $k=6$ we have $-1+\zeta_6=\zeta_3$ and $\zeta_3$ and $-m+\zeta_6$ are always multiplicatively 
dependent.


\section{Canonical number systems and the proof of Theorem \ref{th:bases}}

The main tool in our proof is the following result due to Peth{\H o}
\cite{pethoe1991:polynomial_transformation_and}.

\begin{lemma}[Peth\H{o} {\cite[Theorem 7.1]{pethoe1991:polynomial_transformation_and}}]\label{lem:pethoe}
  Let $P=\sum_{j=0}^dp_jX^j\in\Z[X]$ with $p_d=1$. If $0<p_{d-1}\leq
  p_{d-2}\leq\cdots\leq p_0$, $p_0\geq2$ and no root of $P$ is a root
  of unity, then the pair $(P,\mathcal{N})$ is a canonical number
  system.
\end{lemma}

Since $\zeta_k$ is a primitive $k$-th root of unity $-m+\zeta_k$ is a
root of $P(X)=\Phi_k(m+X)$. By Vieta's formula we obtain
\begin{equation}\label{eq:coeff} p_{\phi(k)-d}=\tilde
  p_d=\sum_{a_1<\cdots<a_d\in(\Z/k\Z)^*}(m-\zeta_k^{a_1})\cdots(m-\zeta_k^{a_d}).
\end{equation}
Here we identify elements of $\Z/k\Z$ with their minimal
representatives in $\Z$, i.e. with the integers in the set
$\{0,1,\dots, k-1\}$ in the obvious way.  In order to apply Lemma
\ref{lem:pethoe} we have to prove that $\tilde p_d\leq\tilde
p_{d+1}$. For a given ordered $(d+1)$-tuple $a_1<\cdots<a_{d+1}$ there
are $d+1$ possibilities to deduce an ordered $d$-tuple. Thus
\begin{multline*}(d+1)\sum_{a_1<\cdots<a_{d+1}\in(\Z/k\Z)^*}(m-\zeta_k^{a_1})\cdots(m-\zeta_k^{a_{d+1}})\\
=\sum_{a_1<\cdots<a_{d}\in(\Z/k\Z)^*}(m-\zeta_k^{a_1})\cdots(m-\zeta_k^{a_{d}})\sum_{\substack{b\in(\Z/k\Z)^*\\b\neq 
a_1,\ldots,a_d}}(m-\zeta_k^b)
\end{multline*}

Therefore we get for $d<\phi(k)$ that
\begin{align*}
\tilde p_{d+1}
=&\sum_{a_1<\cdots<a_{d+1}\in(\Z/k\Z)^*}(m-\zeta_k^{a_1})\cdots(m-\zeta_k^{a_{d+1}})\\
=&\frac{1}{d+1}\sum_{a_1<\cdots<a_{d}\in(\Z/k\Z)^*}(m-\zeta_k^{a_1})\cdots(m-\zeta_k^{a_{d}})\sum_{\substack{b\in(\Z/k\Z)^*\\ b\neq 
a_1,\ldots,a_d}}(m-\zeta_k^b)\\
=&\frac{m(\phi(k)-d)}{d+1}\sum_{a_1<\cdots<a_{d}\in(\Z/k\Z)^*}(m-\zeta_k^{a_1})\cdots(m-\zeta_k^{a_{d}})\\
&-\frac{1}{d+1}\sum_{a_1<\cdots<a_{d}\in(\Z/k\Z)^*}(m-\zeta_k^{a_1})\cdots(m-\zeta_k^{a_{d}})\sum_{\substack{b\in(\Z/k\Z)^*\\ b\neq 
a_1,\ldots,a_d}}\zeta_k^b\\
\geq& \frac{\tilde p_dm(\phi(k)-d)}{d+1}\\
&\quad-\frac{1}{d+1}\left|\sum_{a_1<\cdots<a_{d}\in(\Z/k\Z)^*}(m-\zeta_k^{a_1})\cdots(m-\zeta_k^{a_{d}})\sum_{\substack{b\in(\Z/k\Z)^*\\
b\neq a_1,\ldots,a_d}}\zeta_k^b \right|\\
\geq & \frac{\tilde p_dm(\phi(k)-d)}{d+1}-{\phi(k) \choose {d}}\frac{(m+1)^d(\phi(k)-d)}{d+1}.
\end{align*}
Note that in the first four lines in the inequality above all the sums
over complex terms yield indeed real numbers. Further the ``$\geq$''
signs are true due to the triangle inequality and trivial estimates of
the form $m-1\leq|m-\zeta_k^a|\leq m+1$. Furthermore, by
\eqref{eq:coeff} we have
\[(m+1)^d{\phi(k) \choose {d}}>\tilde p_d>(m-1)^d{\phi(k) \choose
  {d}}\] and therefore
\[\binom{\phi(k)}{d}\frac{(m+1)^d}{\tilde p_d}<\left(\frac{m+1}{m-1}\right)^d=\left(1+\frac{2}{m-1}\right)^d<e^2,\]
where we used that $m-1\geq\phi(k)>d$ by our assumption. Altogether we
obtain the inequality
\[\tilde p_{d+1}\geq \frac{\tilde p_d(m-e^2)(\phi(k)-d)}{d+1}.\]
Thus the coefficients are increasing if 
\begin{equation}\label{eq:bound-m-d}
\frac{(m-e^2)(\phi(k)-d)}{d+1}\geq1.
\end{equation}

We split our considerations into two cases whether $d\leq\phi(k)-2$ or not.
\begin{itemize}
\item\textbf{Case 1: $d\leq\phi(k)-2$.} Together with our assumption
  that $m\geq\phi(k)+1$ we get that the coefficients $\tilde p_d$ up
  to $\tilde p_{\phi(k)-1}$ are increasing
  if $$\frac{2(\phi(k)+1-e^2)}{\phi(k)-1}\geq1.$$ Since $\phi(k)$ only
  takes positive values a simple calculation yields that this is the
  case for $\phi(k)\geq 2e^2-3\approx 11.78$.

  On the contrary, since $\phi(k)$ only takes even values except the
  value $1$, we may assume $\phi(k)\leq 10$, which implies
  $d\leq8$. Plugging this into \eqref{eq:bound-m-d} yields that the
  coefficients are increasing provided $m\geq e^2+\frac92\approx
  11.89$.

  Finally computing all polynomials $\Phi_k(m+x)$ with $\phi(k)+1\leq
  m\leq 11$ by aid of a computer algebra system we see that in any
  case the coefficients $\tilde p_d$ are increasing. In fact we only
  have to check 92 polynomials, which are certainly to many for
  writing them down.
\item\textbf{Case 2: $d=\phi(k)-1$.} Thus we are left to show that
  $p_1\geq p_0$, i.e.
\[\prod_{a \in (\Z/k\Z)^*}(m-\zeta_k^a) \geq \sum_{b \in (\Z/k\Z)^*}\prod_{\substack{a \in (\Z/k\Z)^*\\ a\neq b}}(m-\zeta_k^a).\]
Since
\[1\geq \frac{\phi(k)}{m-1} = \sum_{b \in (\Z/k\Z)^*}\frac{1}{m-1}\geq \sum_{b \in (\Z/k\Z)^*}\frac{1}{m-\zeta_k^b} \]
provided that $m\geq\phi(k)+1$, this case is settled too.
\end{itemize}

The next requirement, in order to apply Lemma \ref{lem:pethoe}, is to
show that no root of $\Phi_k(m+X)$ is a root of unity. Since we assume
$m\geq \phi(k)+1\geq 3$. So all roots of $\Phi_k(m+X)$ have real part $\geq 2$ and therefore cannot be roots of unity.

Finally we have to check that $p_0\geq 2$. But by
\eqref{eq:coeff} we have $p_0\geq (m-1)^{\phi(k)}\geq 2$ provided $m\geq
3$, which we assume.

Therefore $\Phi_k(X+m)$ satisfies all assumptions of Lemma
\ref{lem:pethoe} and Theorem \ref{th:bases} is proved.

\begin{remark}
 In case that $m\leq \phi(k)-1$ we have $p_0<p_1$ and an application of Lemma \ref{lem:pethoe} is no longer possible.
 Indeed, we have
 \[1\leq \frac{\phi(k)}{m+1} = \sum_{b \in (\Z/k\Z)^*}\frac{1}{m+1}< \sum_{b \in (\Z/k\Z)^*}\frac{1}{m-\zeta_k^b} \]
 and by the same arguments as in Case 2 above we obtain $p_0<p_1$. In case that $m=\phi(k)$ both situations $p_1\leq p_0$ and $p_1>p_0$ occur. For 
instance, in case that $m=11$ we have $p_1\leq p_0$ but in case that $m=12$ we have $p_1<p_0$. Therefore in view of the method of proof Theorem \ref{th:bases} 
is best possible.
\end{remark}


\section{Multiplicative independence}

Before we start with the proof of Theorem \ref{th:bases}, let us note that if $p$ and $q$ have a
common factor $d$ then by taking $d$-th roots on both sides of equation \eqref{eq:mult_independent} we obtain
\begin{equation}\label{eq:mult_ind2}
 \zeta_k^j(-m+\zeta_k)^p=(-n+\zeta_k)^q,
\end{equation}
with  $\gcd(p,q)=1$ and $0\leq j<k$.

We start with the proof of the first statement of Theorem
\ref{th:mult_independent}. In view of the discussion below Theorem \ref{th:bases} we consider the Diophantine
equation
\begin{equation}\label{eq:Supper_elliptic} f(x)=y^q,\end{equation}
where $f$ is a polynomial with rational coefficients and with at least
two distinct roots. Obviously \eqref{eq:Norm} fulfills these
requirements as soon as $k>2$.  Due to Schinzel and Tijdeman
\cite{Schinzel:1976} (see also \cite[Theorem 10.2]{Shorey:Dioph}) we
know that for all solutions $(x,y,q)$ to \eqref{eq:Supper_elliptic}
with $|y|>1$, $q$ is bounded by an effectively computable constant
depending only on $f$. In view of Theorem \ref{th:mult_independent} we
see that all solutions coming from~\eqref{eq:mult_independent} yield
$y>1$ provided $n\neq 0$. But it is also well known (see
e.g. \cite[Theorem 6.1 and 6.2]{Shorey:Dioph}) that for given $q\geq
2$ all solutions $(x,y)$ to \eqref{eq:Supper_elliptic} satisfy
$\max\{|x|,|y|\}<C_{f,q}$, where $C_{f,q}$ is an absolute, effectively
computable constant depending on $f$ and $q$. Thus we conclude that
$q=1$ in equation \eqref{eq:mult_ind2} if $m,n>C(k)$, and $C(k)$ is an
effectively computable constant only depending on $k$. Exchanging the
roles of $m$ and $n$ we deduce that $q=p=1$ if $m,n>C(k)$. But this
yields the equation $\zeta_k^j(-m+\zeta_k)=(-n+\zeta_k)$ and taking
norms on both sides gives $\Phi_k(m)=\Phi_k(n)$. Since $\Phi_k(m)$ is
strictly increasing for $m$ large enough we deduce that $m=n$ which we
excluded. Therefore the proof of the first part of
Theorem~\ref{th:mult_independent} is complete.

However, computing the constant $C(k)$ using the methods cited above
will yield very huge constants even for small values of $k$. Therefore
we will use a more direct approach in the proof of the second part.

We note that
$\Phi_n(X)=\Phi_{n/q}(X^{n/q})$ if $q^2|n$. In particular, if $n=q^\ell$ we have
\begin{equation}\label{eq:PowerRed}
\Phi_{q^{\ell}}(X)=\Phi_q\left(X^{q^{\ell -1}}\right). 
\end{equation}
Let us start with powers of $2$. In this case we have that
$\Phi_{2^\ell}(X)=X^{2^{\ell -1}}+1$. Therefore the case that $k$ is a
power of $2$ is deduced from the fact that the Diophantine equation
$$X^2+1=Y^q$$
has no solution. This is a special case of Catalan's equation
$X^p-Y^q=1$ which was completely solved by Mih{\v{a}}ilescu
\cite{Mihailescu:2004}. Note that this special case was already proved
by Chao \cite{Chao:1965}.

In the next step we consider the case that $k$ is a power of $3$ or $6$.

\begin{lemma}\label{lem:k=3}
Let $k$ be a power of $3$ or $6$. Then the two algebraic integers $-m+\zeta_k$ and 
$-n+\zeta_k$ are multiplicatively independent provided $n\neq m$ and $n,m\not\in\{0,-1\}$.
\end{lemma}

\begin{proof}
  Let us consider equation \eqref{eq:Norm} and let $M=m^{k/3}$ if
  $9|k$, $M=m^{k/6}$ if $36|k$ and $M=m$ otherwise. Then equation
  \eqref{eq:Norm} turns into
  \[M^2\pm M+1=\left(\frac{2M\pm 1}2\right)^2+\frac 34=Y^q,\] where
  the ``$+$'' sign holds in case that $k$ is a power of $3$ and the
  ``$-$'' sign holds in case that $k$ is a power of $6$.  Multiplying
  this equation by $4$ and putting $X=2M\pm 1$ we get
  \[X^2+3=4Y^q.\] Due to Luca and Tengely \cite{Luca:2009} we know
  that $(X,Y,q)=(1,1,q)$ and $(37,7,3)$ are the only solutions with
  $X,Y,q\geq 1$. The first solution yields $m\in\{0,-1\}$ which we
  have excluded and the second solution yields $M=\pm 18$ or $M=\pm
  19$. Since both, $18$ and $19$, are not perfect powers we conclude
  that $k=3$ or $k=6$.  We may recover $m$ and $n$ from the
  substitutions that led to equation \eqref{eq:Norm}. Therefore we are
  left to check that in case that $k=3$ equation
  \eqref{eq:mult_independent} has no solution with $m=18,-19$,
  $n=2,-3$, $q=3$ and $p=1$ and in case that $k=6$ equation
  \eqref{eq:mult_independent} has no solution with $m=-18,19$,
  $n=-2,3$, $q=3$ and $p=1$. This is of course easily done e.g. by a
  computer.
\end{proof}

We are left to the case that $k$ is a power of $5,7,11,13,17,19$ or
$23$. Since every solution to \eqref{eq:mult_independent} implies a
solution to the Nagell-Ljunggren equation \eqref{eq:Nagell}, the
following results due to Bugeaud, Hanrot and Mignotte \cite[Theorem 1
and 2]{Bugeaud:2002} are essential in the proof of Theorem
\ref{th:bases}.

\begin{theorem}\label{th:Bugeaud}
The Nagell-Ljunggren equation \eqref{eq:Nagell} has except the three solutions $(x,y,k,q)=(3,11,5,2),(7,20,4,2),(18,7,3,3)$ no further solution in case that
 \begin{itemize}
  \item $k$ is a multiple of $5,7,11$ or $13$, or
  \item $k$ is a multiple of $17,19$ or $23$ and $q\neq 17,19$ or $23$ respectively.
 \end{itemize}
\end{theorem}

Since the identity \eqref{eq:PowerRed}, we get in case that $k=P^\ell$ is a prime power
\[\Phi_{k}(X)=\frac{\left(X^{P^{\ell-1}}\right)^P -1}{\left(X^{P^{\ell-1}}\right)-1}=\frac{x^P-1}{x-1}\]
with $x=X^{P^{\ell -1}}$. If we assume now that $P=5,7,11$ or $13$ the only possible
solutions to \eqref{eq:mult_ind2} imply that $q=1$. Exchanging the role of
$m$ and $n$ we also obtain $p=1$ and equation \eqref{eq:mult_ind2}
turns into
$$\zeta_k^j(-m+\zeta_k)=-n+\zeta_k.$$
Taking norms yields
$$\frac{m^k-1}{m^{k/q}-1}=\frac{n^k-1}{n^{k/q}-1},$$
hence $m=n$, which we excluded. Therefore the we are left to the case that $k$ is a power of $17,19$ or $23$.

Before we consider the remaining cases we want to remind the reader
that an integral basis for $\Z[\zeta_k]$ is given by
$\{1,\zeta_k,\ldots,\zeta_k^{\phi(k)-1}\}$ and therefore every
algebraic integer $\alpha\in\Z[\zeta_k]$ has a unique representation
of the form
\begin{equation}\label{eq:Rep}
\alpha=\sum_{i=0}^{\phi(k)-1} a_i\zeta_k^i, 
\end{equation}
with $a_i\in\Z$ for $0\leq i<\phi(k)$. In the following we will expand
the left and right side of equation \eqref{eq:mult_ind2} in various
cases and bring the result into the unique form \eqref{eq:Rep}. We will
show below that on the left and right side of \eqref{eq:mult_ind2} the
unique representations \eqref{eq:Rep} do not match and conclude that
no solution to \eqref{eq:mult_ind2} and hence no solution to
\eqref{eq:mult_independent} exists.

Assume that \eqref{eq:mult_ind2} has a solution $(m,n,p,q,j)$. In case that $k$ is a power of $17,19$ or $23$ Theorem \ref{th:Bugeaud} implies that the only 
prime divisors of $p$ and $q$ in equation \eqref{eq:mult_independent} are $17,19$ 
or $23$ respectively. Therefore we are left to equation \eqref{eq:mult_ind2}, with $p=1$ and $q=17,19$ or $23$. Note that in case of $k=q$ we have
\begin{align*}
(-n+\zeta_k)^k=&\sum_{i=0}^{k} (-n)^{k-i}{k \choose i} \zeta_k^{i}\\
=&\sum_{i=0}^{k-2} \left((-n)^{k-i}{k \choose i}+kn\right)\zeta_k^i +1=\sum_{i=0}^{k-2} p_{k,i}(n)\zeta_k^i.
\end{align*}

Let us assume that $0\leq j< k-2$. Then with $p=1$ and $k=q$, either the coefficients $p_{k,0}(n)$ and 
$p_{k,1}(n)$ have to 
vanish simultaneously or the coefficients  $p_{k,3}(n)$ and $p_{k,4}(n)$ vanish simultaneously. Since
\[\gcd(p_{17,0}(n),p_{17,1}(n))=\gcd(-n^{17} + 17 n + 1, 17 n^{16} + 17 n)=1 \]
as polynomials in $n$, the coefficients cannot vanish simultaneously. Similarly we obtain
\begin{gather*}
\gcd(p_{19,0}(n),p_{19,1}(n))=\gcd(p_{23,0}(n),p_{23,1}(n))=1\\
\gcd(p_{17,3}(n),p_{17,4}(n))=17n, \quad \gcd(p_{19,3}(n),p_{19,4}(n))=19n,\\
\gcd(p_{23,3}(n),p_{23,4}(n))=23n,
\end{gather*}
where we consider the $\gcd$ taken over the polynomial ring $\Z[n]$. Therefore we deduce $n=0$, a solution which we excluded.

Now consider the case that $k=q$ and $j=k-2$. In this case equation \eqref{eq:mult_ind2} turns into
\[-(m+1)-\zeta_k-\cdots-\zeta_k^{k-2}=(-n+\zeta_k)^k=\sum_{i=0}^{k-2} p_{k,i}(n)\zeta_k^i.\]
Considering the coefficient of $\zeta_k$ we have
\[-1=-kn(1+n^{k-2})\]
which cannot hold if $n>1$.
Similarly for $k=j-1$ we obtain
\[(m+1)+m\zeta_k+\cdots+m\zeta_k^{k-2}=(-n+\zeta_k)^k=\sum_{i=0}^{k-2} p_{k,i}(n)\zeta_k^i.\]
And in particular the coefficients of $\zeta_k$ and $\zeta_k^2$ have to be equal, i.e.
\[-nk-kn^{k-1}=\frac{k(k-1)n^2}2 -kn^{k-1},\]
or equivalently
\[0=nk(nk-n+2).\]
The last equation can only hold if $n$ or $k$ is zero or $nk-n+2=0$ which implies $k=0$ and $n=2$. Since all these options are excluded this case cannot 
occur.

Now let us consider the case that $k=q^\ell$, with $\ell>1$, and $q=17,19$ or $23$. If $j<\phi(k)-1$ in \eqref{eq:mult_ind2} then all coefficients $\zeta_k^i$ 
with $0\leq i <\phi(k)$ of $\zeta_k^j(-m+\zeta_k)$ vanish except exactly two. Indeed we have
\[\zeta_k^j(-m+\zeta_k)=-m\zeta_k^j+\zeta_k^{j+1}.\]
On the right side of \eqref{eq:mult_ind2} all the coefficients of $\zeta_k^i$ with $0\leq i \leq q$ do not vanish, i.e. a contradiction.

In case that $j=\phi(k)+r$ for some $0\leq r < k-\phi(k)-1=q^{\ell -1}-1$ we have
\begin{multline*}
(-m+\zeta_k)\zeta_k^j=(-m+\zeta_k)\left(-\zeta_k^r-\zeta_k^{r+q^{\ell-1}}-\zeta_k^{r+2q^{\ell-1}}-\cdots-\zeta_k^{r+(q-2)q^{\ell-1}}\right)\\
=-m\zeta_k^r +\zeta_k^{r+1}-\cdots -m\zeta_k^{r+(q-2)q^{\ell-1}}+\zeta_k^{r+1+(q-2)q^{\ell-1}}
\end{multline*}
i.e. there exist non vanishing coefficients of $\zeta_k^i$ with $i>q$. But on the right side of \eqref{eq:mult_ind2} all coefficients of $\zeta_k^i$ 
vanish for $i>q$ and again no solution is possible.

We are left to the two cases that $j=\phi(k)-1$ and $j=k-1$. In both cases it is easy to see that $\zeta_k^j(-m+\zeta_k)$ has non vanishing coefficients of 
$\zeta_k^i$ for $i>q$ and we deduce again a contradiction.

\section*{Acknowledgment}
%The authors want to thank Shigeki Akiyama (University of Tsukuba) for
%fruitful discussions on this problem. 

The authors want to thank Shigeki Akiyama (University of Tsukuba) for
pointing them to the refinement of Peth{\H o} \cite[Theorem
7.1]{pethoe1991:polynomial_transformation_and}.

\bibliographystyle{abbrv}
\bibliography{Bases}

\end{document}